---
title: "03: Data manipulation with dplyr"
author: "Brenna Kelly"
format:
  html:
    toc: true
    toc-location: left
editor: source
editor_options: 
  chunk_output_type: console
---

## Base R

We've seen some of the capabilities of what's called `base` R. It can be used for many of the tasks we'll cover in this workshop, and it's highly extensive. We'll now learn from a few **packages** developed to make our lives easier. Because R is open source, anyone can contribute to it — folks can write functions, create a unifying package, and share it with others. (Those that meet certain standards are included in CRAN — the Comprehensive R Archive Network.)

Base R is ready to use as soon as you open a session in R studio. To work with packages (and the handy functions they contain), we need to take a couple extra steps.

## Getting started with a package

The first step is to install a package. This can be done using the GUI or with a line or two or code. Let's first install `dplyr`.

```{r}

# install.packages("dplyr")

```

You only need to install it once (but be sure to update regularly). Once installed, you can load the package with the function `library()`.

```{r}

library(dplyr)

```

Now you're ready to work with the package. To learn more about its functions, you can use `?package_name`. Some packages have better documentation than others, but tidyverse's is considered excellent (this include `dplyr`; more on that later). If you have installed a package but haven't loaded it with `library()` yet, you can still find it using `??package_name` or `??function` (helpful when wondering "what package did that function belong to again?").  

::: callout-tip
Hours of trial and error can save you minutes of reading documentation.
:::

Run the following code to install the packages we'll use in the workshop:  

```{r}

install.packages(c("tidyverse", # note this is a meta-package
                   
                   "RColorBrewer", "colorspace", 
                   "wesanderson")) # because why not


```